---
title: "Join Data with dplyr"
---

```{r setup}
library(tidyverse)
library(nycflights13)

band <- tribble(
   ~name,     ~band,
  "Mick",  "Stones",
  "John", "Beatles",
  "Paul", "Beatles"
)

instrument <- tribble(
    ~name,   ~plays,
   "John", "guitar",
   "Paul",   "bass",
  "Keith", "guitar"
)

instrument2 <- tribble(
    ~artist,   ~plays,
   "John", "guitar",
   "Paul",   "bass",
  "Keith", "guitar"
)
```

# nycflights13

```{r}
View(flights)
```

```{r}
View(airlines)
```

# mutating joins

```{r}
band %>% left_join(instrument, by = "name")
band %>% right_join(instrument, by = "name")
band %>% full_join(instrument, by = "name")
band %>% inner_join(instrument, by = "name")
```

name is the thing in common
if John was capitalized in one and not the other, r wouldn't be able to join it 

## left join 
keeps all the info in the left and only matching info on the right this is why you have NA
## right join
keeps everything on the right and only matching on the left
## full join 
keeps all info from both left and right
## inner join
only keeps matching info

## Your Turn 1a

Which airlines had the largest arrival delays? Complete the code below.

1. Join `airlines` to `flights`


*(Hint: Be sure to remove each `_` before running the code)*

```{r}
flights %>%
  filter(!is.na(arr_delay)) %>%
  left_join(airlines, by = "carrier")  %>%
  head() 
```

## Your Turn 1b

Which airlines had the largest arrival delays? Complete the code below.

1. Join `airlines` to `flights`
2. Compute and order the average arrival delays by airline. Display full names, no codes.

*(Hint: Be sure to remove each `_` before running the code)*

```{r}
flights %>%
  filter(!is.na(arr_delay)) %>%
  left_join(airlines, by = "carrier")  %>%
  group_by(name) %>%
  summarise(delay = mean(arr_delay))  %>%
  arrange(delay) 
```

## Your Turn 2a

Join `flights` and `airports` by `dest` and `faa`. 


```{r}
flights %>% 
  filter(!is.na(arr_delay)) %>%
  left_join(airports, by = c("dest" = "faa")) %>%
  head()

```


## Your Turn 2b

Join `flights` and `airports` by `dest` and `faa`. 

Then for each `name`, compute the `distance` from NYC and the average `arr_delay`.  *Hint: use `first()` to get the first value of distance.*

Order by average delay, worst to best.

*(Hint: Be sure to remove each `_` before running the code)*

```{r}
flights %>% 
  filter(!is.na(arr_delay)) %>%
  left_join(airports, by = c("dest" = "faa")) %>%
  group_by(name) %>%
  summarise(distance = first(distance), delay = mean(arr_delay)) %>%
  arrange(desc(delay))

```



# filtering joins

```{r}
band %>% semi_join(instrument, by = "name")
band %>% anti_join(instrument, by = "name")
```
## semi
only matching values from the left that had a match in the right
## anti
keeps only what did not match from the left in the right
## Your Turn 3

How many airports in `airports` are serviced by flights in `flights`? (i.e. how many places can you fly to direct from New York?) 

Notice that the column to join on is named `faa` in the **airports** data set and `dest` in the **flights** data set.

```{r}
airports %>%
  semi_join(flights, by = c("faa" = "dest")) %>%
  select(faa)
```


# other functions  

```{r}
distinct(instrument, plays)
```


# Take aways

* `left_join()` retains all cases in _left_ data set
* `right_join()` retains all cases in _right_ data set
* `full_join()` retains all cases in _either_ data set
* `inner_join()` retains _only_ cases in _both_ data sets

* `semi_join()` extracts cases that _have_ a match
* `anti_join()` extracts cases that _do not have_ a match 


